Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2018-08-27 article.Rnw
Make changes in the paper (improves and corrects the text)

% -------------------------------------------
% Abstract
\Abstract{ In the analysis of count data often the equidispersion
  assumption is not suitable, hence the Poisson regression model is
  inappropriate.  As a generalization of the Poisson distribution the
  COM-Poisson distribution can deal with under-, equi- and overdispersed
  count data.  It is a member of the exponential family of distributions
  and has the Poisson and geometric distributions as special cases, as
  well as the Bernoulli distribution as a limiting case.  In spite of
  the nice properties of the COM-Poisson distribution, its location
  parameter does not correspond to the expectation, which complicates
  the interpretation of regression models specified using this
  distribution. In this paper, we propose a straightforward
  reparametrization of the COM-Poisson distribution based on an
  approximation to the expectation of this distribution. The main
  advantage of our new parametrization is the straightforward
  interpretation of the regression coefficients in terms of the
  expectation of the count response variable, as usual in the context of
  generalized linear models. Furthermore, the estimation and inference
  for the new COM-Poisson regression model can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likeli-hood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. We observed that the empirical correlation
  between the regression and dispersion parameter estimators is close to
  zero, which suggests that these parameters are orthogonal. We
  illustrate the application of the proposed model through the analysis
  of three data sets with over-, under- and equidispersed count data.
  The study of distribution properties through a consideration of
  dispersion, zero-inflated and heavy tail indexes, together with the
  results of data analysis show the flexibility over standard
  approaches. Therefore, we encourage the application of the new
  parametrization for the analysis of count data in the context of
  COM-Poisson regression models.  The computational routines for fitting
  the original and new version of the COM-Poisson regression model and
  the analyzed data sets are available in the supplementary material.
}

@@

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
on the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such a restriction can be unsuitable, since the
observed data can present  variance both smaller or larger than the
expectation, leading to the cases of under and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non-equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

@@

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Specfically for the
overdispersion case Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where
the expectation of the Poisson distribution is assumed to be gamma
distributed.  However, other distributions can be used to represent the
random variation.  For example the Poisson-Tweedie
model~\citep{Bonat2017} and its special cases as the Poisson
inverse-Gaussian and Neyman-Type A assume that the random effects are
Tweedie, inverse Gaussian and Poisson distributed, respectively.  The
Gamma-Count distribution assumes a gamma distribution for the time
between events, thus it can deal with underdispersed as well as
overdispersed count data~\citep{Zeviani2014}. The COM-Poisson
distribution is obtained by a generalization of the Poisson distribution
allowing for a non-linear decrease in the ratios of successive
probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution is a member of the exponential family and
it has the Poisson and geometric distributions as special cases, as well
as the Bernoulli distribution as a limiting case. It can deal with both
under and overdispersed count data.  Some recently applications of the
COM-Poisson distribution include \citet{Lord2010} for the analysis of
traffic crash data, \citet{Sellers2010} for the modelling of airfreight
breakage and book purchases, and \citet{Huang2017} to the analysis of
attendance data, takeover birds and cotton boll counts.  The main
disadvantage of the COM-Poisson regression model as presented
in~\citet{Sellers2010} is that its location parameter does not
correspond to the expectation of the distribution, which complicates the
interpretation of regression models and means that they are not
comparable with standard approaches such as the Poisson and negative
binomial regression models. \citet{Huang2017} proposed a
mean-parametrization of the COM-Poisson distribution in order to avoid
such an issue.  In this approach the mean parameter is obtained by
solving an non-linear equation defined as an infinite sum.
Consequently, it is computationally demanding and liable to numerical
problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
expectation and $\phi$ is a dispersion parameter. In contrast to the
original parametrization, the proposed parametrization leads to
regression coefficients directly associated with the expectation of the
response variable, as usual in the context of generalized linear models.
Consequently, the obtained regression coefficients are comparable with
the ones obtained by standard approaches, such as the Poisson and
negative binomial regression models.  Furthermore, our novel COM-Poisson
parametrization is simpler than the strategy proposed
by~\cite{Huang2017}, since it does not require any numerical method for
solving non-linear equations, and we show the attractive properties like
the orthogonality between dispersion and regression parameters and
consistency and asymptotic normality of the maximum likelihood
estimators are retained.

@@

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using the original parametrization. In this case, the COM-Poisson
regression model is $\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the
relationship between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable parametrization. In the
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} proposes that the
parameter $\lambda$ as a function of $\mu$ and $\nu$, is given by the
solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(y!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression model is $\log(\mu_i)
= \bm{x}_i^\top \bm{\beta}$. In this article, we propose an alternative
mean-parametrization of the COM-Poisson distribution in order to avoid the
limitations of the original parametrization and the numerical complexity
of the \Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on
the mean approximation (\autoref{eqn:mean-approx}). We introduced a new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h(\lambda, \nu) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h^{-1}(\mu, \nu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but on another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as a special case.

@@

In order to assess the accuracy of the moment approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for (a) expectation  and (b) variance.  The
quadratic errors were obtained by $[\mu - \text{E}(Y)]^2$ for the
expectation and by $[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the
variance. In both cases $\text{E}(Y)$ and Var$(Y)$ were computed
numerically. The dotted lines represent the border between the regions
$\nu \leq 1$ and $\lambda > 10^\nu$, in the $\mu$ and $\phi$ scale.

@@

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly dependent
on the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribution is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicate the distribution is in general a
light-tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.

@@

Parameter estimation requires the numerical maximization of
\autoref{eqn:ll-rcmp}. Since the derivatives of $\ell$ cannot be
obtained in closed forms, we adopted the \texttt{BFGS}
algorithm~\citep{Nocedal1995} as implemented in the function
\texttt{optim()} for the statistical software \texttt{R}
\citep{Rcore2017}. Standard errors for the regression coefficients are
obtained based on the observed information matrix
$\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ (hessian matrix)
is computed numerically. Confidence intervals for $\hat{\mu}_i$ are
obtained by using the delta method~\citep[p. 89]{Pawitan2001}.

@@

We considered average counts varying from $3$ to $27$ according to a
regression model with a continuous and a categorical covariate. The
continuous covariate~($\bm{x}_1$) was generated as a sequence from $0$
to $1$ and of length equal to the sample size.  Similarly, the categorical
covariate~($\bm{x}_2$) was generated as a sequence of three values each
one repeated $n/3$ times (rounding up when required), where $n$ denotes
the sample size. Thus, the expectation of the COM-Poisson random
variable is given by
$\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21} \bm{x}_{21} +
\beta_{22} \bm{x}_{22})$,
where $\bm{x}_{21}$ and $\bm{x}_{22}$ are dummy representing the levels
of $\bm{x}_2$.  The regression coefficients were fixed at the values,
$\beta_0 = 2$, $\beta_1 = 0.5$, $\beta_{21} = 0.8$ and
$\beta_{22} = -0.8$.

@@

%-----------------------------------------------------------------------
% Explain the simulation
% \begin{algorithm}
% \caption{Steps in simulation study.}
% \label{alg:simulation}
%   \Begin{
%   $\bm{\beta} = \begin{bmatrix} \beta_0 & \beta_1 & \beta_{21} &
%     \beta_{22} \end{bmatrix}^\top = \begin{bmatrix} 2.0 &
%     0.5 & 0.8 & -0.8 \end{bmatrix}^\top$\;
%   \For{$n \in \{50, 100, 300, 1000\}$}{
%     set $\bm{x}_1$ as a sequence, with $n$ elements, between $0$ and
%     $1$\;
%     set $\bm{x}_2$ as a repetition, with $n$ elements, of three
%     categories\;
%     compute $\bm{\mu}$ using $\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1
%     +  \beta_{21} \bm{x}_{21} + \beta_{22} \bm{x}_{22})$, where
%     $\bm{x}_{21}$ and $\bm{x}_{22}$ are \textit{dummy} variable for
%     $\bm{x}_2$\;
%     \For{$\phi \in \{-1.6, -1.0, 0.0, 1.8\}$}{
%       \Repeat{$1000$ times}{
%         simulate $\bm{y}$ from COM-Poisson distribution with $\bm{\mu}$
%         and $\phi$ parameters\;
%         fit COM-Poisson$_\mu$ regression model to $\bm{y}$ data with
%         $\bm{X} = \begin{bmatrix} \bm{1} & \bm{x}_1 & \bm{x}_{21} &
%           \bm{x}_{22} \end{bmatrix}$ design matrix\;
%         get $\hat{\bm{\theta}} = \begin{bmatrix} \hat{\phi} & \hat{\beta}_0 &
%           \hat{\beta}_1 & \hat{\beta}_{21} & \hat{\beta}_{22}
%         \end{bmatrix}^\top$\;
%         get confidence intervals for $\hat{\bm{\theta}}$ by quadratic
%         approximation at the maximum likelihood estimate
%         (assumes $\hat{\bm{\theta}} \sim \mathcal{N}(\hat{\bm{\theta}},
%         \sqrt{-\bm{v}})$, where $\bm{v}$ is a diagonal of the inverse of
%         the Hessian matrix)\;
%       }
%     }
%   }
% }
% \end{algorithm}

@@

In order to check the consistency of the estimators we considered four
different sample sizes: $50$, $100$, $300$ and $1000$; generating $1000$
data sets in each case.  In \autoref{fig:bias-plot}, we show the bias of
the estimators for each simulation scenario (combination between values
of the dispersion parameter and samples sizes) along with the confidence
intervals calculated as average bias plus and minus $\Phi(0.975)$ times
the average standard error. The scales are standardized for each
parameter by dividing the average bias by the average standard error
obtained for the sample of size $50$.

@@

To illustrate the orthogonality, \autoref{fig:ortho-surf} displays
contour plots of the deviance surfaces for four simulated data set of
size 1000 with $\mu=5$ and different values of the dispersion
parameters. The shapes of the deviance function show that the proposed
parametrization is better for both computation and asymptotic
(normal-based) inference. Furthermore, it is interesting to note that
the deviance function shape under strong overdispersion ($\phi=-1.6$) is
not as well behaved as the others; this is due to the difficulty of the
distribution in dealing with strong overdispersion in low counts (see
dispersion index plot in the \autoref{fig:indexes-plot}). This also
explains the results of Cov$(\hat{\beta}_{22}, \phi)$ in the first panel
of \autoref{fig:ortho-plot}, since $\beta_{22}$ is negative and
associated with low counts.

@@

Following~\citet{Zeviani2014}, the linear predictor is given by
$$\log(\mu_{ij}) = \beta_0 + \beta_{1j} \texttt{def}_i + \beta_{2j}
\texttt{def}_i^2$$
where $\mu_{ij}$ is the expected number of cotton bolls for the $i$-th
defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5: 100\%)
and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3: blossom,
4: boll, 5: boll open), i.e. we have a second order effect of
defoliation in each growth stage. The parameters estimates and
goodness-of-fit measures for the Poisson, COM-Poisson, COM-Poisson$_\mu$
and quasi-Poisson regression models are presented in
\autoref{tab:coef-cotton}.

@@

\begin{table}[ht]
\centering
\caption{Empirical correlations between $\hat{\phi}$ and
  $\hat{\bm{\beta}}$ for the two parametrizations of COM-Poisson model
  fit to underdispersed data.}
\label{tab:corr-cotton}
\begingroup\small
\begin{tabular}{lrrrrrr}
  \toprule
<<cres-coton, results="asis">>=
@
\end{tabular}
\endgroup
\end{table}

@@

For the analysis of this data set based on the descriptive analysis
(\autoref{fig:desc-soy}), we proposed the following linear predictor
$$
\log(\mu_{ijk}) = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
where $i=$1: block II, 2: block III, 3: block IV e 4: block V; $j=$1:
50\% e 2: 62.5\%; and $k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg
dm$^{-3}$, where $\gamma_i$ is the effect of $i$-th block ($i=$1: block
II, 2: block III, 3: block IV and 4: block V), $\tau_j$ is the effect of
$j$-th moisture level ($j=$1: 50\% and 2: 62.5\%) and $\beta_{3j}$ is
interaction of the first order potassium effect (\texttt{K}) for the
$j$-th moisture level (\texttt{umid}). \autoref{tab:coef-soy} presents
the estimates, ratio between estimate and standard error and
goodness-of-fit measures for the alternative models.

@@

For this data set we consider three models with linear predictors,
\begin{center}
\begin{minipage}{12cm}
Linear: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$\\
Quadratic: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Cubic: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}

@@

In this paper, we presented and characterized a novel parametrization of
the COM-Poisson distribution and the associated regression model. The
novel parametrization was based on a simple asymptotic approximation for
the expectation and variance of the COM-Poisson distribution. The main
advantage of the proposed reparametrization is the simple interpretation
of the regression coefficients in terms of the expectation of the
response variable as usual in the generalized linear models
context. Thus, it is possible to compare the results of the COM-Poisson
model with the ones from standard approaches as the Poisson and
quasi-Poisson regression models. Furthermore, in the novel
parametrization the COM-Poisson distribution is indexed by the
expectation $\mu$ and an extra dispersion parameter $\phi$ which our
data analysis suggest being orthogonal. This is similar to
\Citeauthor{Huang2017}'s (\citeyear{Huang2017}) parametrization but is
simpler, because the $\mu$ is obtained from simple algebra.

We evaluated the accuracy of the asymptotic approximations for the
expectation and variance of the COM-Poisson distribution by considering
quadratic approximation errors.  The results showed that the
approximations are accurate for a large part of the parameter space,
which in turn support our reparametrization. Moreover, we discuss the
properties and flexibility of the distribution to deal with count data
although dispersion, zero-inflation and heavy-tail indexes. We carried
out a simulation study to assess the properties of the reparametrized
COM-Poisson model to deal with different levels of dispersion as well as
the properties of the maximum likelihood estimators. The results of our
simulation study suggested that the maximum likelihood estimators of the
regression and dispersion parameters are unbiased and consistent.  The
empirical coverage rates of the confidence intervals computed based on
the asymptotic distribution of the maximum likelihood estimators are
close to the nominal level for sample size greater than $100$.  The
worst scenario is when we have small sample sizes and strong
overdispersed counts.  In general, we recommend the use of the
asymptotic confidence intervals for computational simplicity.

>>>>

% -------------------------------------------
% Abstract
\Abstract{The COM-Poisson distribution is a two-parameter generalization
  of Poisson distribution can deal with under-, equi- and overdispersed
  count data. Its location parameter does not correspond to the
  expectation, which complicates the interpretation of regression models
  specified using this distribution. In this paper, we propose a
  straightforward reparametrization of the COM-Poisson distribution
  based on an approximation to the expectation of this distribution. The
  main advantage of our new parametrization is the interpretation of the
  regression coefficients in terms of the expectation of the count
  response variable, as usual in the context of generalized linear
  models. Furthermore, the estimation and inference for the
  reparametrized COM-Poisson regression models can be done based on the
  likelihood paradigm. We carried out simulation studies to verify the
  finite sample properties of the maximum likelihood estimators. The
  results from our simulation study show that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  dispersion parameters. The empirical correlation between the
  regression and dispersion parameter estimators and deviance surfaces
  suggests that these parameters are orthogonal. We illustrate the
  application of the proposed model through the analysis of three data
  sets with over-, under- and equidispersed count data. The study of
  distribution properties through a consideration of dispersion,
  zero-inflated and heavy tail indexes, together with the results of
  data analysis show the flexibility over standard approaches.  The
  computational routines for fitting the original and new version of the
  COM-Poisson regression model and the analyzed data sets are available
  in the supplementary material.  }

@@

In spite of the advantages of the Poisson regression model, the Poisson
distribution has only one parameter, which represents both the
expectation and variance of the count random variable.  This restriction
on the relationship between the expectation and variance induced by the
Poisson distribution is referred as equidispersion.  However, in
practical data analysis such restriction can be unsuitable, since the
observed data can present  variance both smaller or larger than the
expectation, leading to the cases of under- and overdispersion,
respectively.  The main problem of the application of the Poisson
regression model to non-equidispersed count data is that the standard
errors associated with the regression coefficients are inconsistently
estimated, which in turn can lead to misleading
inferences~\citep{Winkelmann1995, Bonat2017}.

@@

The strategies for constructing alternative count distributions are
related with the causes of the non-equidispersion. Specifically for the
overdispersion, Poisson mixture models are widely applied.  One
popular example of this approach is the negative-binomial model, where
the expectation of the Poisson distribution is assumed to be gamma
distributed.  However, other distributions can be used to represent the
random variation.  For example the Poisson-Tweedie
model~\citep{Bonat2017} and its special cases as the Poisson
inverse-Gaussian and Neyman-Type A assume that the random effects are
Tweedie, inverse Gaussian and Poisson distributed, respectively.  The
Gamma-Count distribution assumes a gamma distribution for the time
between events, thus it can deal with underdispersed as well as
overdispersed count data~\citep{Zeviani2014}. The COM-Poisson
distribution is obtained by a generalization of the Poisson distribution
allowing for a non-linear decrease in the ratios of successive
probabilities~\citep{Shmueli2005}.

The COM-Poisson distribution has the Poisson and geometric distributions
as special cases, as well as the Bernoulli distribution as a limiting
case. It can deal with both under- and overdispersed count data.  Some
recently applications of the COM-Poisson distribution include
\citet{Lord2010} for the analysis of traffic crash data,
\citet{Sellers2010} for the modelling of airfreight breakage and book
purchases, \citet{Huang2017} to the analysis of attendance data,
takeover birds and cotton boll counts, and \citet{Chatla2018} to model
counts from bike sharing systems. Theoretical results and approximations
derived from this distribution is discussed by \citet{Shmueli2005},
\citet{Daly2016} and \citet{Gaunt2017}. The main disadvantage of the
COM-Poisson regression model as presented in~\citet{Sellers2010} is that
its location parameter does not correspond to the expectation of the
distribution, which complicates the interpretation of regression models
and means that they are not comparable with standard approaches such as
the Poisson and negative binomial regression models. \citet{Huang2017}
proposed a mean-parametrization of the COM-Poisson distribution in order
to avoid such an issue.  In this approach the mean parameter is obtained
by solving an non-linear equation defined as an infinite sum.
Consequently, it is computationally demanding and liable to numerical
problems.

The main goal of this article is to propose a novel COM-Poisson
parametrization based on the mean approximation presented by
\citet{Shmueli2005}. In this parametrization, the probability mass
function is written in terms of $\mu$ and $\phi$, where $\mu$ is the
approximated expectation and $\phi$ is a dispersion parameter. In
contrast to the original parametrization, the proposed parametrization
leads to regression coefficients directly associated with the
expectation of the response variable, as usual in the context of
generalized linear models.  Consequently, the obtained regression
coefficients are comparable with the ones obtained by standard
approaches, such as the Poisson and negative binomial regression models.
Furthermore, our novel COM-Poisson parametrization is simpler than the
strategy proposed by~\cite{Huang2017}, since it does not require any
numerical method for solving non-linear equations, and we show the
attractive properties like the orthogonality between dispersion and
regression parameters and consistency and asymptotic normality of the
maximum likelihood estimators are retained.

@@

The COM-Poisson regression model was proposed by~\citet{Sellers2010},
using the original parametrization. In this case, the COM-Poisson
regression model is $\log(\lambda_i) = \bm{x}_i^\top \bm{\beta}$ and the
relationship between E$(Y_i)$ and $\bm{x}_i$ is modelled indirectly.
\citet{Huang2017} shows how to model directly the expectation of the
COM-Poisson distribution in a suitable parametrization. In the
\autoref{eqn:pmf-cmp}, \Citeauthor{Huang2017} proposes that the
parameter $\lambda$ as a function of $\mu$ and $\nu$, is given by the
solution to
\begin{equation*}
  \sum_{j=0}^{\infty} (j - \mu) \frac{\lambda^j}{(j!)^\nu} = 0\,.
\end{equation*}
Thus the mean-parametrized COM-Poisson regression model is $\log(\mu_i)
= \bm{x}_i^\top \bm{\beta}$. In this article, we propose an alternative
mean-parametrization of the COM-Poisson distribution in order to avoid the
limitations of the original parametrization and the numerical complexity
of the \Citeauthor{Huang2017}'s approach.

@@

The proposed reparametrization of COM-Poisson models is based on
the mean approximation (\autoref{eqn:mean-approx}). We introduced a new
parameter $\mu$, using this approximation,
\begin{equation}
  \label{eqn:repar-cmp}
  \mu = h_\nu(\lambda) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
  \quad \Rightarrow \quad
  \lambda = h_\nu^{-1}(\mu) = \left (\mu +
    \frac{(\nu - 1)}{2\nu} \right )^\nu.
\end{equation}
The dispersion parameter is taken on the log scale for computational
convenience, thus $\phi = \log(\nu)$, $\phi \in \mathbb{R}$. The
interpretation of $\phi$ is the same as the $\nu$, but on another
scale. For $\phi < 0$ and $\phi > 0$ we have the overdispersed and
underdispersed cases, respectively.  When $\phi=0$ we have Poisson
distribution as a special case.

@@

In order to assess the accuracy of the moment approximations
(\autoref{eqn:mean-approx}), \autoref{fig:approx-plot} presents the
quadratic errors for (a) expectation and (b) variance.  The quadratic
errors were obtained by $[\mu - \text{E}(Y)]^2$ for the expectation and
by $[ \mu \exp(-\phi) - \text{Var}(Y)]^2$ for the variance. In both
cases $\text{E}(Y)$ and Var$(Y)$ were computed numerically\footnote{We
  compute E$(Y) = \sum_{y=0}^{500}yp(y)$ and
  E$(Y^2) = \sum_{y=0}^{500}y^2p(y)$.}. The dotted lines represent the
border between the regions $\nu \leq 1$ and $\lambda > 10^\nu$, in the
$\mu$ and $\phi$ scale.

@@

The constraint $\mu>0$ imposes an undesirable restriction in the
original parameter space, $\lambda^\nu>(\nu-1)/(2\nu)$. However, this
restricted region is related to small underdispersed counts (averages
smaller than 0.1 and $\nu>1$) a parameter region unexplored in practice.

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly dependent
on the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribution is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicate the distribution is in general a
light-tailed distribution, i.e. $\text{HT} \to 0$ for $y \to \infty$.

@@

Parameter estimation requires the numerical maximization of
\autoref{eqn:ll-rcmp}. Since the derivatives of $\ell$ cannot be
obtained in closed forms, we obtain the Hessian matrix by central finite
differences. We adopted the \texttt{BFGS} algorithm~\citep{Nocedal1995}
as implemented in the function \texttt{optim()} for the statistical
software \texttt{R} \citep{Rcore2017}. Standard errors for the
regression coefficients are obtained based on the observed information
matrix $\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ (Hessian matrix)
is computed numerically. Confidence intervals for $\hat{\mu}_i$ are
obtained by using the delta method~\citep[p. 89]{Pawitan2001}.

@@

We considered average counts varying from $3$ to $27$ according to a
regression model with a continuous and a categorical covariate. The
continuous covariate~($\bm{x}_1$) was generated as a sequence from $0$
to $1$ and of length equal to the sample size.  Similarly, the
categorical covariate~($\bm{x}_2$) was generated as a sequence of three
values each one repeated $n/3$ times (rounding up when required), where
$n$ denotes the sample size. The parameter $\mu$ of the reparametrized
COM-Poisson random variable is given by
$\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21} \bm{x}_{21} +
\beta_{22} \bm{x}_{22})$,
where $\bm{x}_{21}$ and $\bm{x}_{22}$ are dummy representing the levels of
$\bm{x}_2$.  The regression coefficients were fixed at the values,
$\beta_0 = 2$, $\beta_1 = 0.5$, $\beta_{21} = 0.8$ and
$\beta_{22} = -0.8$.

@@

@@

In order to check the consistency of the estimators we considered four
different sample sizes: $50$, $100$, $300$ and $1000$; generating $1000$
data sets in each case. For sample sizes 50 and 100, we have 29 and 8
simulations of the 1000 where algorithm did not converge. These non
convergence situations occurred for $\phi=-1.6$. In
\autoref{fig:bias-plot}, we show the bias of the estimators for each
simulation scenario (combination between values of the dispersion
parameter and samples sizes) along with the confidence intervals
calculated as average bias plus and minus $\Phi(0.975)$ times the
average standard error. The scales are standardized for each parameter
by dividing the average bias by the average standard error obtained for
the sample of size $50$.

@@

To illustrate the orthogonality, \autoref{fig:ortho-surf} displays
contour plots of the deviance surfaces for four simulated data set of
size 1000 with $\mu=5$ and different values of the dispersion
parameters. The shapes of the deviance function show that the proposed
parametrization is better for both computation and asymptotic
(normal-based) inference. Furthermore, it is interesting to note that
the deviance function shape under strong overdispersion ($\phi=-1.6$) is
not as well behaved as the others; this is due to the accurary of mean
approximation \autoref{fig:approx-plot}. Although the reparametrized
model is always valid, $\mu$ and $\phi$ are orthogonal only when $\mu$
is the expectation of the distribution, in other words, when
approximation $\lambda - (\nu-1)/(2\nu)$ is accurate. This also explains
the results of Cov$(\hat{\beta}_{22}, \phi)$ in the first panel of
\autoref{fig:ortho-plot}, since $\beta_{22}$ is negative and associated
with low counts.

@@

Following~\citet{Zeviani2014}, the linear predictor is given by
$$\eta_i = \beta_0 + \beta_{1j} \texttt{def}_i + \beta_{2j}
\texttt{def}_i^2$$
where $\eta_i = \log(\lambda_i)$ for the original COM-Poisson and
$\eta_i = \log(\mu_{ij})$ for COM-Poisson$_\mu$. The parameter $\mu_i$
is the (approximated) expected number of cotton bolls for the $i$-th
defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5: 100\%)
and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3: blossom,
4: boll, 5: boll open), i.e. we have a second order effect of
defoliation in each growth stage. The parameters estimates and
goodness-of-fit measures for the Poisson, COM-Poisson, COM-Poisson$_\mu$
and quasi-Poisson regression models are presented in
\autoref{tab:coef-cotton}.

@@

\begin{table}[ht]
\centering
\caption{Empirical correlations between $\hat{\phi}$ and
  $\hat{\bm{\beta}}$ for the two parametrizations of COM-Poisson model
  fit to underdispersed data.}
\label{tab:corr-cotton}
\begingroup\small
\begin{tabular}{lrrrrrr}
  \toprule
<<cres-cotton, results="asis">>=
@
\end{tabular}
\endgroup
\end{table}

@@

For the analysis of this data set based on the descriptive analysis
(\autoref{fig:desc-soy}), we proposed the following linear predictor
$$
\eta_i = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
where $i=$1: block II, 2: block III, 3: block IV e 4: block V; $j=$1:
50\% e 2: 62.5\%; and $k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg
dm$^{-3}$, where $\gamma_i$ is the effect of $i$-th block ($i=$1: block
II, 2: block III, 3: block IV and 4: block V), $\tau_j$ is the effect of
$j$-th moisture level ($j=$1: 50\% and 2: 62.5\%) and $\beta_{3j}$ is
interaction of the first order potassium effect (\texttt{K}) for the
$j$-th moisture level (\texttt{umid}); The predictor $\eta_i$ is
$\log(\lambda_i)$ for the original COM-Poisson and $\log(\mu_{ij})$ for
COM-Poisson$_\mu$.\autoref{tab:coef-soy} presents the estimates, ratio
between estimate and standard error and goodness-of-fit measures for the
alternative models.

@@

For this data set we consider three models with linear predictors,
\begin{center}
\begin{minipage}{12cm}
Linear: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i$\\
Quadratic: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Cubic: $\eta_i = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}
where $\eta_i = \log(\lambda_i)$ for the original COM-Poisson and
$\eta_i = \log(\mu_{ij})$ for COM-Poisson$_\mu$.

@@

In this paper, we presented and characterized a novel parametrization of
the COM-Poisson distribution and the associated regression model. The
novel parametrization was based on a simple asymptotic approximation for
the expectation of the COM-Poisson distribution. The main advantage of
the proposed reparametrization is the simple interpretation of the
regression coefficients in terms of the expectation of the response
variable as usual in the generalized linear models context. Thus, it is
possible to compare the results of the COM-Poisson model with the ones
from standard approaches as the Poisson and quasi-Poisson regression
models. Furthermore, in the novel parametrization the COM-Poisson
distribution is indexed by $\mu$ and an extra dispersion parameter
$\phi$ which our data analysis suggest being orthogonal. This is similar
to \Citeauthor{Huang2017}'s (\citeyear{Huang2017}) parametrization but
is simpler, because the $\mu$ is obtained from simple algebra.

The proposed parametrization is always valid, only the interpretation of
$\mu$ parameter, as expectation of the distribution, depends on the
accuracy of the approximation. We evaluated the accuracy of the
approximations for the expectation and variance of the COM-Poisson
distribution by considering quadratic approximation errors.  The results
showed that the approximation for the expectation is accurate for a
large part of the parameter space, which in turn support our
reparametrization. Moreover, we discuss the properties and flexibility
of the distribution to deal with count data although dispersion,
zero-inflation and heavy-tail indexes. We carried out a simulation study
to assess the properties of the reparametrized COM-Poisson model to deal
with different levels of dispersion as well as the properties of the
maximum likelihood estimators. The results of our simulation study
suggested that the maximum likelihood estimators of the regression and
dispersion parameters are unbiased and consistent.  The empirical
coverage rates of the confidence intervals computed based on the
asymptotic distribution of the maximum likelihood estimators are close
to the nominal level for sample size greater than $100$.  The worst
scenario is when we have small sample sizes and strong overdispersed
counts.  In general, we recommend the use of the asymptotic confidence
intervals for computational simplicity.
